---
title: "je2 - analysis"
author: "BrunoPalau"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(tidyverse)
library(tidyr)
library(stats)
library(ggpubr)
library(ggalluvial)
library(readxl)
```

import data
```{r}
df <- read_excel("../../Data/SA_evolution_phenotypes_v2_noCowan.xlsx")
set.seed(123)
```

Data transformation
```{r}
df <- df%>%
  filter(SA_strain_ID == "JE2")

# only select feature columns
features <- df %>%
  select("growth_in_supernatant", "growth_in_TSB", "average_STX", "average_H2O2_survival","Hemolysis_score")

# log transform all relative features
features$log_average_STX <- log(features$average_STX)
features$log_average_H2O2_survival <- log(features$average_H2O2_survival)
# features$log_Hemolysis_score <- features$Hemolysis_score

# normalize values
features <- as.data.frame(sapply(colnames(features),function(c){
  x <- as.matrix(features)[,c]
  return ((2 * ( (x-min(x)) / (max(x)- min(x)) ) ) - 1)
})
)

# keep og hemolysis score
features$org_hemolysis_score <- as.character(df$Hemolysis_score)
features$org_growth_in_supernatant <- df$growth_in_supernatant
features$org_growth_in_TSB <- df$growth_in_TSB
features$org_STX <- df$average_STX
features$org_H2O2 <- df$average_H2O2_survival



# perform pca
pcas <- prcomp(features[,c("growth_in_supernatant", "growth_in_TSB", "log_average_STX", "log_average_H2O2_survival","Hemolysis_score")])

# store in features dataframe
features$PC1 <- pcas$x[,1]
features$PC2 <- pcas$x[,2]
features$PC3 <- pcas$x[,3]
features$PC4 <- pcas$x[,4]
features$Strain <- df$SA_strain_ID
features$Condition <- df$evolution_condition
features$ID <- df$strain_clone_ID
```

pc visualization

````{r}
# plot on 2 pc
ggplot(features,
       aes(x = PC1, y= PC2,color=Condition))+
  geom_point()

```


compute optimal number of clusters
```{r}
library(cluster)
library(factoextra)

set.seed(123)

# Elbow method
fviz_nbclust(features[,c("growth_in_supernatant", "growth_in_TSB", "log_average_STX", "log_average_H2O2_survival","Hemolysis_score")], kmeans, method = "wss") +
    geom_vline(xintercept = 4, linetype = 2)+
  labs(subtitle = "Elbow method")

# Silhouette method
fviz_nbclust(features[,c("growth_in_supernatant", "growth_in_TSB", "log_average_STX", "log_average_H2O2_survival","Hemolysis_score")], kmeans, method = "silhouette")+
  labs(subtitle = "Silhouette method")

# Gap statistic
# nboot = 50 to keep the function speedy. 
# recommended value: nboot= 500 for your analysis.
# Use verbose = FALSE to hide computing progression.
set.seed(123)
fviz_nbclust(features[,c("growth_in_supernatant", "growth_in_TSB", "log_average_STX", "log_average_H2O2_survival","Hemolysis_score")], kmeans, nstart = 25,  method = "gap_stat", nboot = 50)+
  labs(subtitle = "Gap statistic method")


nr_clusters <- 4
```

cluster with optimal nr. of clusters
```{r}
set.seed(150)
# cluster on features
features$phenotype_clusters = as.factor(kmeans(features[,c("growth_in_supernatant", "growth_in_TSB", "log_average_STX", "log_average_H2O2_survival","Hemolysis_score")],nr_clusters)$cluster)
```

visualize clusters
```{r}
ggplot(features,
       aes(x = PC1, y= PC2,color=Condition,shape=phenotype_clusters))+
  geom_point()
```

proportion of condition found in each cluster
```{r}
ggplot(features, aes(x = phenotype_clusters, fill=Condition))+
  geom_bar()
```

get the id of wrong clustering TSB clones
```{r}
tsb_wrong_clustering <- features%>%
  filter(Condition == "TSB" & phenotype_clusters == 1) %>%
  select(ID, Condition)

supernatant_wrong_clustering <- features%>%
  filter(Condition == "supernatant" & phenotype_clusters == 1) %>%
  select(ID, Condition)

print(tsb_wrong_clustering)
print(supernatant_wrong_clustering)
```


```{r}
ggplot(features%>%group_by(phenotype_clusters), aes(x = phenotype_clusters, y= log_average_STX))+
  geom_boxplot()

ggplot(features%>%group_by(phenotype_clusters), aes(x = phenotype_clusters, y= growth_in_TSB))+
  geom_boxplot()

ggplot(features%>%group_by(phenotype_clusters), aes(x = phenotype_clusters, y= growth_in_supernatant))+
  geom_boxplot()

ggplot(features%>%group_by(phenotype_clusters), aes(x = phenotype_clusters, y= log_average_H2O2_survival))+
  geom_boxplot()
```

complex heatmap visualization
```{r}
library(ComplexHeatmap)
library(circlize)

pdf("../Out/JE2/heatmap_JE2.pdf", width = 16, height = 8)

# seperate relative and absolute values
stx_max = max(abs(min(log(features$org_STX))),max(log(features$org_STX)))
h2o2_max = max(abs(min(log(features$org_H2O2))),max(log(features$org_H2O2)))

relative_m = t(as.matrix(features%>%
                           mutate(log_STX_production = log(org_STX)/stx_max,
                                  log_H2O2_survival = log(org_H2O2)/h2o2_max) %>%
                           select(log_STX_production,log_H2O2_survival)))

absolute_m = t(as.matrix(features%>%select(org_growth_in_supernatant,org_growth_in_TSB)))

colnames(relative_m) <- features$ID
colnames(absolute_m) <- features$ID

# choose colors
col_fun_relative = colorRamp2(c(-1,0,1), c("DeepSkyBlue","white","purple"))
col_fun_absolute= colorRamp2(c(min(absolute_m),(min(absolute_m) + max(absolute_m))/2,max(absolute_m)), c("white","red","darkred"))

# column annotations
cluster_ha = columnAnnotation(Phenotype_cluster = features$phenotype_clusters, name = "Clusters by Phenotype", col = list(Phenotype_cluster =  c("1" = "Turquoise", "2" = "Gold", "3" = "CornflowerBlue", "4" = "FireBrick")))

condition_ha = columnAnnotation(Evolution_condition = features$Condition, name = "Evolution Condition", col = list(Evolution_condition =  c("supernatant" = "OrangeRed", "TSB" = "Chartreuse")))

hemolysis_ha = HeatmapAnnotation(Hemolysis_score = as.factor(features$org_hemolysis_score), name = "Hemolysis score",
    col = list(Hemolysis_score =  c("0" = "orange", "1" = "yellow", "2"= "green")),
    annotation_legend_param = list(Hemolysis_score = list(labels = c("0 - absent", "1 - reduced", "2 - wild type")))
    )


relative_hm <- Heatmap(relative_m,top_annotation = c(condition_ha,cluster_ha),
                           #column_km = nr_clusters,
        #column_split = features$Condition,
        #column_gap = unit(5, "mm"),
        heatmap_legend_param = list(title = "normalized values", labels = c("-1 - less than ancestor","-0.5"," 0 - same as ancestor","0.5","1 - more than ancestor")),
        col = col_fun_relative,
        show_row_dend = FALSE,
        show_column_dend = TRUE,
        # cluster_columns = FALSE,
        column_split = features$phenotype_clusters,
        border = TRUE
)

aboslute_hm <- Heatmap(absolute_m,
        bottom_annotation = hemolysis_ha,
        #column_split = features$Condition,
        #column_gap = unit(5, "mm"),
        heatmap_legend_param = list(title = "absolute growth"),
        col = col_fun_absolute,
        show_row_dend = FALSE,
        show_column_dend = FALSE,
        # cluster_columns = FALSE,
        border = TRUE
)


je2_hm <- relative_hm %v% aboslute_hm

draw(je2_hm)
dev.off()
```




# cluster only on one feature

supernatant growth clusters
```{r}
features$supernatant_growth_clusters = as.factor(kmeans(features[,c("growth_in_supernatant")],nr_clusters)$cluster)


ggplot(features,
       aes(x = PC1, y= PC2,color=supernatant_growth_clusters,shape=Condition))+
  geom_point()

```

```{r}
ggplot(features, aes(x = supernatant_growth_clusters, fill=Condition))+
  geom_bar()
```

tsb growth cluster
```{r}
features$TSB_growth_clusters = as.factor(kmeans(features[,c("growth_in_TSB")],nr_clusters)$cluster)


ggplot(features,
       aes(x = PC1, y= PC2,color=TSB_growth_clusters,shape=Condition))+
  geom_point()

```


```{r}
ggplot(features, aes(x = TSB_growth_clusters, fill=Condition))+
  geom_bar()
```

STX cluster
```{r}
features$STX_clusters = as.factor(kmeans(features[,c("average_STX")],nr_clusters)$cluster)


ggplot(features,
       aes(x = PC1, y= PC2,color=STX_clusters,shape=Condition))+
  geom_point()

```

```{r}
ggplot(features, aes(x = STX_clusters, fill=Condition))+
  geom_bar()
```


log STX cluster
```{r}
features$log_STX_clusters = as.factor(kmeans(features[,c("log_average_STX")],nr_clusters)$cluster)


ggplot(features,
       aes(x = PC1, y= PC2,color=log_STX_clusters,shape=Condition))+
  geom_point()

```

```{r}
ggplot(features, aes(x = log_STX_clusters, fill=Condition))+
  geom_bar()
```




log H2O2 clusters
```{r}
features$log_H2O2_clusters = as.factor(kmeans(features[,c("log_average_H2O2_survival")],nr_clusters)$cluster)


ggplot(features,
       aes(x = PC1, y= PC2,color=log_H2O2_clusters,shape=Condition))+
  geom_point()

```

```{r}
ggplot(features, aes(x = log_H2O2_clusters, fill=Condition))+
  geom_bar()
```

visualize hemolysis distribution on condition
```{r}
ggplot(features,
       aes(x = PC1, y= PC2,color=as.factor(org_hemolysis_score),shape=Condition))+
  geom_point()

```

```{r}
ggplot(features, aes(x = org_hemolysis_score, fill=Condition))+
  geom_bar()
```

save changes
```{r}
write.csv(features%>%select(c("ID","Strain","Condition","phenotype_clusters", "PC1", "PC2")),"../Out/JE2/cluster_results_JE2.csv")
```


store condition clusters
```{r}
features <- features%>%
  mutate(cond_clusters = ifelse(Condition == "TSB",1,2))


write.csv(features%>%
            # filter(phenotype_clusters_2 == 1 | Condition == "supernatant")%>%
            select(c("ID","Strain","Condition","cond_clusters","PC1", "PC2")),"../Out/JE2/condition_results_JE2.csv")
```

```{r}
ggplot(features, aes(x = log_average_STX, y = log_average_H2O2_survival, color = Condition)) +
geom_point() +
  geom_smooth(method = "lm")
```

```{r}
ggplot(features, aes(x = log_average_STX, y = org_growth_in_supernatant)) +
geom_point() +
  geom_smooth()
```